---
title: "Project 3"
output:
  html_document:
    theme:
      version: 4
---

```{r setup, include=FALSE}
library(tidyverse)
library(colorspace)
knitr::opts_chunk$set(echo = TRUE)
```

This is the dataset you will be working with:

```{r message = FALSE, warning = FALSE}
food <- readr::read_csv("https://wilkelab.org/DSC385/datasets/food_coded.csv")
food
```

A detailed data dictionary for this dataset is available [here.](https://wilkelab.org/DSC385/datasets/food_codebook.pdf) The dataset was originally downloaded from Kaggle, and you can find additional information about the dataset [here.](https://www.kaggle.com/datasets/borapajo/food-choices)


**Question:** Is GPA related to student income, the mother's educational level, or the student's relationship status?

To answer this question, first prepare a cleaned dataset that contains only the four relevant data columns, properly cleaned so that numerical values are stored as numbers and categorical values are represented by humanly readable words or phrases. For categorical variables with an inherent order, make sure the levels are in the correct order.

In your introduction, carefully describe each of the four relevant data columns. In your analysis, provide a summary of each of the four columns, using `summary()` for numerical variables and `table()` for categorical variables.

Then, make one visualization each for student income, mother's educational level, and relationship status (column `marital_status`), and answer the question separately for each visualization. The three visualizations can be of the same type.

**Hints:**

1. Use `case_when()` to recode categorical variables.

2. Use `fct_relevel()` to arrange categorical variables in the right order.

3. Use `as.numeric()` to convert character strings into numerical values. It is fine to ignore warnings about `NA`s introduced by coercion.

4. `NaN` stands for Not a Number and can be treated like `NA`. You do not need to replace `NaN` with `NA`.

5. When using `table()`, provide the argument `useNA = "ifany"` to make sure missing values are counted: `table(..., useNA = "ifany")`.

**Introduction:** <br />
This analysis aims to explore whether there is a relationship between students’ GPA and three other factors: their income, their mother’s educational level, and their relationship status. We focus on four relevant columns from the dataset:<br />

**1. GPA (gpa):** This column represents the Grade Point Average of students, a numerical value typically ranging between 2.0 and 4.0, where higher values reflect better academic performance. The GPA will be treated as a continuous numeric variable.<br />

**2. Income (income):** This column records students' self-reported income levels, with categorical values representing different income brackets. This variable will be treated as a categorical factor, and missing (NA) values will also be considered in the analysis.<br />

**3. Mother’s Educational Level (mother_education):** This column indicates the highest level of education attained by the student’s mother. The values are categorical, ordered from lowest (e.g., "Did not complete high school") to highest educational attainment (e.g., "Graduate degree"). Since there is a natural order, this variable will be treated as an ordinal categorical factor.<br />

**4. Relationship Status (marital_status):** This column shows the student’s relationship status (e.g., "Single," "In a relationship," "Married"). These values are unordered categorical variables, and we will treat this variable as a nominal factor. Missing (NA) values will also be handled and highlighted in the visualizations.

**Approach:** <br />
To investigate the relationship between GPA and each of these three factors, we will clean the dataset and ensure that the variables are properly formatted. The income, mother_education, and marital_status columns will be converted to factors, with appropriate handling of missing values. The GPA column will be treated as a continuous numeric variable.<br />

- We will start by summarizing the four columns using summary statistics for GPA and frequency tables for the categorical variables.
- For each categorical variable (income, mother’s educational level, and relationship status), we will create boxplots to visually assess the distribution of GPA across the different categories. NA values in the categorical variables will be highlighted in gray (if any) in the plots.
- Each visualization will be analyzed separately to explore any potential trends or relationships between GPA and the categorical variables.

**Analysis:**

```{r }

# Clean the dataset
cleaned_food <- food %>%
  select(GPA, income, mother_education, marital_status) %>%
  mutate(
    # Convert GPA to numeric
    gpa = as.numeric(GPA),
    
    # Recode student income into readable categories
    income = case_when(
      income == 1 ~ "Less than $15,000",
      income == 2 ~ "$15,001 to $30,000",
      income == 3 ~ "$30,001 to $50,000",
      income == 4 ~ "$50,001 to $70,000",
      income == 5 ~ "$70,001 to $100,000",
      income == 6 ~ "Higher than $100,000",
      TRUE ~ NA_character_
    ),
    
    # Recode mother_education and ensure proper level ordering
    mother_education = case_when(
      mother_education == 1 ~ "Less than high school",
      mother_education == 2 ~ "High school degree",
      mother_education == 3 ~ "Some college degree",
      mother_education == 4 ~ "College degree",
      mother_education == 5 ~ "Graduate degree",
      TRUE ~ NA_character_
    ),
    mother_education = fct_relevel(mother_education, 
                                   "Less than high school", "High school degree", "Some college degree",
                                   "College degree", "Graduate degree"),
    
    # Recode relationship status
    marital_status = case_when(
      marital_status == 1 ~ "Single",
      marital_status == 2 ~ "In a relationship",
      marital_status == 3 ~ "Cohabiting",
      marital_status == 4 ~ "Married",
      marital_status == 5 ~ "Divorced",
      marital_status == 6 ~ "Widowed",
      TRUE ~ NA_character_
    )
  )

# Summarize numerical variables
summary(cleaned_food$gpa)


# Summarize categorical variables
table(cleaned_food$income, useNA = "ifany")
table(cleaned_food$mother_education, useNA = "ifany")
table(cleaned_food$marital_status, useNA = "ifany")

```

```{r }

# Visualization for Income vs GPA
ggplot(cleaned_food, aes(x = gpa, y = fct_infreq(income), fill =is.na(income))) +
  geom_boxplot(outlier.color = 'blue', outlier.size = 3, show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "gray", "FALSE" = "skyblue")) +
  theme_minimal(12) +
  labs(title = "GPA vs Student Income") +
  scale_x_continuous(
    name = "Students' GPA",
    limits = c(2.0,4.0)
  ) +
  scale_y_discrete(
    name = "Students' Income"
  ) +
  theme(
    axis.text.y = element_text(vjust = 0),
     axis.title.y = element_text(vjust = 3),
    axis.title.x = element_text(vjust = -2),
    plot.title = element_text(hjust = 0.5)
  )

# Visualization for Mother's Education vs GPA
  ggplot(cleaned_food, aes(x = gpa, y = mother_education, fill = is.na(mother_education))) +
  geom_boxplot(outlier.color = 'violetred', outlier.size = 3, show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "gray", "FALSE" = "lightpink")) +  
  theme_minimal(12) +
  labs(title = "GPA vs Mother's Education") +
  scale_x_continuous(
    name = "Students' GPA",
    limits = c(2.0,4.0)
  ) +
  scale_y_discrete(
    name = "Mother's Education"
  ) +
  theme(
    axis.text.y = element_text(vjust = 0),
    axis.title.y = element_text(vjust = 3),
    axis.title.x = element_text(vjust = -2),
    plot.title = element_text(hjust = 0.5)
  ) 

# Visualization for Relationship Status vs GPA
ggplot(cleaned_food, aes(x = gpa, y = fct_infreq(marital_status), fill = is.na(marital_status))) +
  geom_boxplot(outlier.color = 'tomato', outlier.size = 3, show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "gray", "FALSE" = "lightsalmon")) +
  theme_minimal(12) +
  labs(title = "GPA vs Relationship Status") +
  scale_x_continuous(
    name = "Students' GPA",
    limits = c(2.0,4.0)
  ) +
  scale_y_discrete(
    name = "Students' Relationship Status"
  ) +
  theme(
    axis.text.y = element_text(vjust = 0),
    axis.title.y = element_text(vjust = 3),
    axis.title.x = element_text(vjust = -2),
    plot.title = element_text(hjust = 0.5)
  )

```

**Discussion:** <br />
In the **GPA vs Student Income** plot, the distribution of GPA across different income levels shows some variability but no strong, clear trend. Most income brackets have a similar median GPA, centered around 3.3 to 3.6, with slight differences in spread. For example, students in the lower income bracket (*Less than $15,000*) exhibit a wide spread of GPAs, but their median GPA is the same those in middle-income brackets such as *$50,001 to $70,000*. In higher-income groups, such as *$70,001 to $100,000* and *Higher than $100,000*, the spread of GPA values is also wide, and outliers (marked in blue) indicate a few students with significantly lower GPAs compared to their peers within the same income category.

Overall, the plot suggests that while there are minor variations in GPA across different income levels, there is no strong relationship between student income and GPA. The distributions remain fairly consistent across most income brackets, with medians hovering around the same range. The "NA" category for missing income data lacks a boxplot due to insufficient data, further highlighting the lack of information for that group.<br />

In the **GPA vs Mother's Education** plot, we observe a broader range of GPAs for students based on their mother's educational level. Students whose mothers have *"Less than high school"* education exhibit the widest spread of GPA values, ranging from around 2.2 to 3.8, with a median GPA near 2.7. This indicates a wide variability in academic performance for this group.

Two notable outliers are present, represented by magenta dots. One outlier appears in the *"Graduate degree"* category with a GPA of exactly 4.0, indicating a student with perfect academic performance. Another outlier is in the *"College degree"* category, with a GPA slightly below 2.25, which is significantly lower than the rest of the students in this group.

These outliers highlight that while most students' GPAs fall within expected ranges for each educational category, there are some extreme cases that deviate from the norm.

Students whose mothers have a *"Some college degree"* and *"Graduate degree"* have similar medians around 3.4, while the *"Graduate degree"* group tends to have a more compact GPA distribution. Interestingly, the *"High school degree"* group has the highest median among all the groups, around 3.6, indicating more consistent performance among students in this category.

The *"NA"* group represents students with missing data regarding their mother's education. These students have a narrower GPA range and a median lower than the other groups except for the *"Less than high school"* group.

Overall, there is some indication that higher maternal education might be associated with slightly higher and more consistent GPAs, though the differences between categories are not dramatic.<br />

In the **GPA vs. Relationship Status** plot, we observe the distribution of students' GPAs across different relationship statuses. The box plot suggests that students who are either single or in a relationship tend to have higher GPAs, with most data points concentrated around 3.0 to 4.0 GPA. We also observe a few outliers (orange points), highlighted in blue, particularly for students who are single or in a relationship. These outliers represent GPAs that deviate significantly from the main cluster of data. Specifically, there are single and "in a relationship" students with GPAs below 2.5, which are much lower than the typical range observed for these categories. These lower GPAs are exceptions and may indicate unique circumstances or factors affecting those students. While the bulk of students in these categories tend to have higher GPAs (closer to the 3.0-4.0 range), these outliers suggest that relationship status alone may not fully account for academic performance, and other factors might be at play. Nevertheless, the overall trend indicates a stronger performance for single and "in a relationship" students, with a few exceptions as highlighted by the outliers.<br />

On the other hand, the GPA data for married students and those with missing relationship status (NA) show lack of data points, as indicated by the absence of boxplots for these groups. With so few/no data points available, we cannot draw concrete conclusions about the relationship between GPA and these statuses. Overall, it appears that students who are single or in a relationship tend to maintain higher GPAs, but the limited data for married students and those with an unknown status makes it difficult to generalize the overall effect of relationship status on GPA. Collecting more data for these categories could provide clearer insights.

